---
title: "PHP 2517 Homework #1"
author: "Blain Morin"
date: "February 11, 2019"
output: pdf_document
header-includes:
- \usepackage{float}
---

```{r setup, include=FALSE}

### Set knit options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)


### Load required libraries
library(lme4)
library(ggplot2)
library(dplyr)
library(stargazer)
library(readr)
library(extrafont)
library(grid)
library(gridExtra)


### Load Data
cd4 = read_csv("cd4.csv")

### Set Seed
set.seed(121)

```

# GH Chapter 11: Exercise 4

## a.) Graph the outcome (the CD4 percentage, on the square root scale) for 10 children as a function of time.

```{r}

### Get 10 random IDs
ten.kid.ids = cd4 %>%
  select(newpid) %>%
  distinct() %>%
  sample_n(size = 10)

### Get data for 10 IDs
ten.kid.data = cd4 %>%
  filter(newpid %in% as.numeric(as.matrix(ten.kid.ids))) %>%
  mutate(newpid = as.factor(newpid))

### Plot for 10 random individuals
ten.kid.data %>%
  ggplot(aes(x = time, y = sqrt(CD4PCT))) +
  geom_line(aes(group = newpid, color = newpid)) +
  geom_point(aes(color = newpid)) +
  ylab("sqrt(CD4PCT)") +
  xlab("Time") +
  ggtitle("CD4 Trajectories for 10 Random Children") +
  theme_classic() +
  theme(text=element_text(size=12,  family="CM Sans"))
  
  

```

## b.) Each child’s data has a time course that can be summarized by a linear fit. Estimate these lines and plot them for 10 children.

```{r}

### Plot individually fitted LMs
ten.kid.data %>%
  ggplot(aes(x = time, y = sqrt(CD4PCT))) +
  stat_smooth(aes(group = newpid, color = newpid), method = "lm", se = FALSE) +
  geom_point(aes(color = newpid)) +
  ylab("sqrt(CD4PCT)") +
  xlab("Time") +
  ggtitle("Individual LM for 10 Random Children") +
  theme_classic() +
  theme(text=element_text(size=12,  family="CM Sans"))

```

## c.) Set up a model for the children’s slopes and intercepts as a function of the treatment and age at baseline. Estimate this model using the two-step procedure–first estimate the intercept and slope separately for each child, then fit the between-child models using the point estimates from the first step.

```{r, fig.height=8}

### Change treatment to factor
cd4 = cd4 %>%
  mutate(treatmnt = as.factor(treatmnt))

### Function that gets intercepts from LM
get.ints = function(x, y) {
  
  model.i = lm(y ~ x)
  return(coef(model.i)[1])
  
}


### Function that gets slopes from LM
get.slopes = function(x, y) {
  
  model.i = lm(y ~ x)
  return(coef(model.i)[2])
  
}

### Get ints and slopes for each individual regression
cd4 = cd4 %>%
  group_by(newpid) %>%
  mutate(beta0 = get.ints(x = time, y = sqrt(CD4PCT))) %>%
  mutate(beta1 = get.slopes(x = time, y = sqrt(CD4PCT))) %>%
  ungroup()

### Regress intercept on tx and age
b0.model = lm(beta0 ~ treatmnt + baseage, data = cd4)
b0.model1 = lm(beta0 ~ treatmnt + log(baseage), data = cd4)


### REsidual plots for intercept regression
b0.resid.plot = ggplot(b0.model) +
  geom_point(aes(x = .fitted, y = .resid)) +
  xlab("Fitted") +
  ylab("Residual") +
  ggtitle("Residual Plot: Linear Age vs Intercept") +
  theme_classic() +
  theme(text=element_text(size=12,  family="CM Sans"))
  

b0.resid.plot1 = ggplot(b0.model1) +
  geom_point(aes(x = .fitted, y = .resid)) +
  xlab("Fitted") +
  ylab("Residual") +
  ggtitle("Residual Plot: log(Age) vs Intercept") +
  theme_classic() +
  theme(text=element_text(size=12,  family="CM Sans"))



### Regress slope on tx and age
b1.model = lm(beta1 ~ treatmnt + baseage, data = cd4)
b1.model1 = lm(beta1 ~ treatmnt + log(baseage), data = cd4)


### Residual plots for slope
b1.resid.plot = ggplot(b1.model) +
  geom_point(aes(x = .fitted, y = .resid)) +
  xlab("Fitted") +
  ylab("Residual") +
  ggtitle("Residual Plot: Linear Age vs Slope") +
  theme_classic() +
  theme(text=element_text(size=12,  family="CM Sans"))
  

b1.resid.plot1 = ggplot(b1.model1) +
  geom_point(aes(x = .fitted, y = .resid)) +
  xlab("Fitted") +
  ylab("Residual") +
  ggtitle("Residual Plot: log(Age) vs Slope") +
  theme_classic() +
  theme(text=element_text(size=12,  family="CM Sans"))

### Arrange plots
grid.arrange(b0.resid.plot,
             b0.resid.plot1,
             b1.resid.plot,
             b1.resid.plot1,
             nrow = 4)
```

```{r, results = 'asis'}

stargazer(b0.model1, b1.model1,
          header = FALSE,
          table.placement = 'H',
          title = "Regression on Intercept and Slope",
          covariate.labels = c("Treatment = 2",
                               "log(baseage)"),
          notes = "Individuals with only one observation do not have a slope estimate.")

```
